---
title: "PSC_preprocessing_and_celltyping"
author: "Paola Pibiri"
date: "2023-11-08"
output: html_document
---

```{r header, include=FALSE}
###################################################################################################################
# Authors: Paola Pibiri
# Name: PSC_preprocessing_and_celltyping.Rmd
# Function: perform all the steps necessary for the preprocessing and the celltyping in the PSC project
###################################################################################################################
```


## Libraries
```{r}
#############################################
# libraries                                 #
#############################################

library(Seurat)
library(patchwork)
library(ggplot2)
library('UCell')
library(MAST)
```


## Functions
```{r}

plot_ncount_vs_mitopct <- function(seurat_object_metadata){
  # totally stolen from Harm
  p <- ggplot(seurat_object_metadata, aes(nCount_RNA, percent.mt)) + geom_hex(bins=100) +
    scale_fill_distiller(palette = "Spectral", name="Cell frequencies",
                         limits = c(0,100), oob = scales::squish) +
    ylab("Fraction mtDNA-encoded genes") + xlab("Number of UMIs") +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          panel.background = element_blank(), axis.line = element_line(colour = "grey"),
          axis.text=element_text(size=12), axis.title=element_text(size=18))
  return(p)
}

add_data_from_rownames <- function(table_w_conditions, condition_column='condition', split_char='_', column_names=NULL){
  # get the data to split
  data_to_split <- table_w_conditions[[condition_column]]
  # create an emtpy dataframe
  metadata_dataframe <- NULL
  # check each rowname
  for(i in 1 : length(data_to_split)){
    # get the value
    value <- data_to_split[i]
    # split the data
    split_data <- strsplit(value, split_char)
    # initialize the dataframe
    if(is.null(metadata_dataframe)){
      metadata_dataframe <- data.frame(matrix(, nrow=length(data_to_split), ncol = length(split_data[[1]])))
    }
    # add this row
    metadata_dataframe[i, ] <- split_data[[1]]
  }
  # add the column names if requested
  if(!is.null(column_names)){
    colnames(metadata_dataframe) <- column_names
  }
  # add to the original data
  table_w_conditions <- cbind(table_w_conditions, metadata_dataframe)
  return(table_w_conditions)
}


```

## Preprocessing
```{r main}
#read the original object
psc <- readRDS('/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/PSC_2022_integrated_postqc_filtered.rds')

# Preprocessing
DefaultAssay(psc)<-"SCT" 
set.seed(0)
psc <- FindVariableFeatures(psc, assay = 'SCT')
#
psc<- RunPCA(psc)
psc <- RunUMAP(psc, dims = 1:30)
psc<-FindNeighbors(psc, dims = 1:30)
psc<-FindClusters(psc, resolution = 1.2)
psc_plot <- DimPlot(psc, group.by = "seurat_clusters", label = T, label.box = T, pt.size = 0.5, repel = T) + NoLegend()

ggsave("psc_clusters_mt70.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")
saveRDS(psc, "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/psc_clusters.rds")

```

## Division in Compartments: Epithelial
```{r }

#define the main marker
feature_epi_epcam <- FeaturePlot(psc, features=c('EPCAM'))

# put the average expression per cluster into a dataframe
avg_exp_epcam <- AverageExpression(psc, features = "EPCAM", assays = "RNA")
avg_exp_epcam_table <- data.frame(cluster=colnames(avg_exp_epcam[['RNA']]), expression=as.vector(unlist(avg_exp_epcam[['RNA']])))
#
psc@meta.data$epcam_positive <- NA
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_epcam_table[avg_exp_epcam_table[['expression']] > 4, 'cluster'], 'epcam_positive_4'] <- T
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_epcam_table[avg_exp_epcam_table[['expression']] <= 4, 'cluster'], 'epcam_positive_4'] <- F
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_epcam_table[avg_exp_epcam_table[['expression']] > 5, 'cluster'], 'epcam_positive_5'] <- T
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_epcam_table[avg_exp_epcam_table[['expression']] <= 5, 'cluster'], 'epcam_positive_5'] <- F
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_epcam_table[avg_exp_epcam_table[['expression']] > 3, 'cluster'], 'epcam_positive_3'] <- T
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_epcam_table[avg_exp_epcam_table[['expression']] <= 3, 'cluster'], 'epcam_positive_3'] <- F

epi_epcam_4 <- DimPlot(psc, group.by = 'epcam_positive_4') + scale_color_manual(values=c('gray', 'red'))
epi_epcam_5 <- DimPlot(psc, group.by = 'epcam_positive_5') + scale_color_manual(values=c('gray', 'red'))
epi_epcam_3 <- DimPlot(psc, group.by = 'epcam_positive_3') + scale_color_manual(values=c('gray', 'red'))


plot_grid(psc_plot, feature_epi_epcam, epi_epcam_5, epi_epcam_3, epi_epcam_4, nrow = 2, ncol = 3)
ggsave("epi_epcam1.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

```


## Division in Compartments: Stromal
```{r }

#define the main stromal markers
feature_stro_thy1 <- FeaturePlot(psc, features=c('THY1'))
feature_stro_sox10 <- FeaturePlot(psc, features=c('SOX10'))
feature_stro_madcam1 <- FeaturePlot(psc, features=c('MADCAM1'))

##### Check THY marker
avg_exp_thy1 <- AverageExpression(psc, features = "THY1", assays = "RNA")
avg_exp_thy1_table <- data.frame(cluster=colnames(avg_exp_thy1[['RNA']]), expression=as.vector(unlist(avg_exp_thy1[['RNA']])))

psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_thy1_table[avg_exp_thy1_table[['expression']] > 0.4, 'cluster'], 'thy1_positive_0.4'] <- T
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_thy1_table[avg_exp_thy1_table[['expression']] <= 0.4, 'cluster'], 'thy1_positive_0.4'] <- F
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_thy1_table[avg_exp_thy1_table[['expression']] > 1, 'cluster'], 'thy1_positive_1'] <- T
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_thy1_table[avg_exp_thy1_table[['expression']] <= 1, 'cluster'], 'thy1_positive_1'] <- F
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_thy1_table[avg_exp_thy1_table[['expression']] > 0.3, 'cluster'], 'thy1_positive_0.3'] <- T
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_thy1_table[avg_exp_thy1_table[['expression']] <= 0.3, 'cluster'], 'thy1_positive_0.3'] <- F


stro_thy1_0.4 <- DimPlot(psc, group.by = 'thy1_positive_0.4') + scale_color_manual(values=c('gray', 'red'))
stro_thy1_1 <- DimPlot(psc, group.by = 'thy1_positive_1') + scale_color_manual(values=c('gray', 'red'))
stro_thy1_0.3 <- DimPlot(psc, group.by = 'thy1_positive_0.3') + scale_color_manual(values=c('gray', 'red'))


plot_grid(psc_plot, feature_stro_thy1, stro_thy1_0.4, stro_thy1_0.3, stro_thy1_1)
ggsave("stro_thy1.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")


##### Check SOX10 marker
avg_exp_sox10 <- AverageExpression(psc, features = "SOX10", assays = "RNA")
avg_exp_sox10_table <- data.frame(cluster=colnames(avg_exp_sox10[['RNA']]), expression=as.vector(unlist(avg_exp_sox10[['RNA']])))
psc@meta.data$sox10_positive <- NA
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_sox10_table[avg_exp_sox10_table[['expression']] > 2, 'cluster'], 'sox10_positive_2'] <- T
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_sox10_table[avg_exp_sox10_table[['expression']] <= 2, 'cluster'], 'sox10_positive_2'] <- F
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_sox10_table[avg_exp_sox10_table[['expression']] > 2.5, 'cluster'], 'sox10_positive_2.5'] <- T
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_sox10_table[avg_exp_sox10_table[['expression']] <= 2.5, 'cluster'], 'sox10_positive_2.5'] <- F


stro_sox10_2 <- DimPlot(psc, group.by = 'sox10_positive_2') + scale_color_manual(values=c('gray', 'red'))
stro_sox10_2.5 <- DimPlot(psc, group.by = 'sox10_positive_2.5') + scale_color_manual(values=c('gray', 'red'))

plot_grid(psc_plot, feature_stro_sox10, stro_sox10_2, stro_sox10_2.5)
ggsave("stro_sox10.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")


##### Check MADCAM1 marker
avg_exp_madcam1 <- AverageExpression(psc, features = "MADCAM1", assays = "RNA")
avg_exp_madcam1_table <- data.frame(cluster=colnames(avg_exp_madcam1[['RNA']]), expression=as.vector(unlist(avg_exp_madcam1[['RNA']])))

psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_madcam1_table[avg_exp_madcam1_table[['expression']] > 1, 'cluster'], 'madcam1_positive_1'] <- T
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_madcam1_table[avg_exp_madcam1_table[['expression']] <= 1, 'cluster'], 'madcam1_positive_1'] <- F
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_madcam1_table[avg_exp_madcam1_table[['expression']] > 2, 'cluster'], 'madcam1_positive_2'] <- T
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_madcam1_table[avg_exp_madcam1_table[['expression']] <= 2, 'cluster'], 'madcam1_positive_2'] <- F

stro_madcam1_1 <- DimPlot(psc, group.by = 'madcam1_positive_1') + scale_color_manual(values=c('gray', 'red'))
stro_madcam1_2 <- DimPlot(psc, group.by = 'madcam1_positive_2') + scale_color_manual(values=c('gray', 'red'))

plot_grid(psc_plot, feature_stro_madcam1, stro_madcam1_1, stro_madcam1_2)
ggsave("stro_madcam1.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plotss")


```


## Division in Compartments: Immune
```{r }

#define the main markers od the immune compartment
feature_imm_ptprc <- FeaturePlot(psc, features=c('PTPRC'))
feature_imm_cd27 <- FeaturePlot(psc, features=c('CD27'))

##### check the PTPRC marker
avg_exp_ptprc <- AverageExpression(psc, features = "PTPRC", assays = "RNA")
avg_exp_ptprc_table <- data.frame(cluster=colnames(avg_exp_ptprc[['RNA']]), expression=as.vector(unlist(avg_exp_ptprc[['RNA']])))
#psc@meta.data$madcam1_positive <- NA
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_ptprc_table[avg_exp_ptprc_table[['expression']] > 3, 'cluster'], 'ptprc_positive_3'] <- T
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_ptprc_table[avg_exp_ptprc_table[['expression']] <= 3, 'cluster'], 'ptprc_positive_3'] <- F
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_ptprc_table[avg_exp_ptprc_table[['expression']] > 3.5, 'cluster'], 'ptprc_positive_3.5'] <- T
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_ptprc_table[avg_exp_ptprc_table[['expression']] <= 3.5, 'cluster'], 'ptprc_positive_3.5'] <- F

imm_ptprc_3 <- DimPlot(psc, group.by = 'ptprc_positive_3') + scale_color_manual(values=c('gray', 'red'))
imm_ptprc_3.5 <- DimPlot(psc, group.by = 'ptprc_positive_3.5') + scale_color_manual(values=c('gray', 'red'))

plot_grid(psc_plot, feature_imm_ptprc, imm_ptprc_3, imm_ptprc_3.5)
ggsave("imm_ptprc.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")


##### check the CD27 marker
avg_exp_cd27 <- AverageExpression(psc, features = "CD27", assays = "RNA")
avg_exp_cd27_table <- data.frame(cluster=colnames(avg_exp_cd27[['RNA']]), expression=as.vector(unlist(avg_exp_cd27[['RNA']])))


psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_cd27_table[avg_exp_cd27_table[['expression']] >= 0.3, 'cluster'], 'cd27_positive_0.3'] <- T
psc@meta.data[psc@meta.data[['seurat_clusters']] %in% avg_exp_cd27_table[avg_exp_cd27_table[['expression']] < 0.3, 'cluster'], 'cd27_positive_0.3'] <- F

imm_cd27_0.3 <- DimPlot(psc, group.by = 'cd27_positive_0.3') + scale_color_manual(values=c('gray', 'red'))

plot_grid(psc_plot, feature_imm_cd27, imm_cd27_0.3)
ggsave("imm_cd27.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")


```


## Division in Compartments: Final
```{r }

psc@meta.data$compartment <- 'none'
psc@meta.data[psc@meta.data$epcam_positive_4 == T, 'compartment'] <- 'epithelial'
psc@meta.data[psc@meta.data$ptprc_positive_3 == T |
               psc@meta.data$cd27_positive_0.3 == T, 'compartment'] <- 'immune'
psc@meta.data[psc@meta.data$thy1_positive_0.3 == T |
               psc@meta.data$sox10_positive_2 == T |
               psc@meta.data$madcam1_positive_1 == T, 'compartment'] <- 'stromal'

compart <- DimPlot(psc, group.by = 'compartment') + scale_color_manual(values=c('firebrick', 'dodgerblue4', 'green4'))
plot_grid(psc_plot, compart)

ggsave("psc_compartments2.pdf", width = 20, height = 16, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")
saveRDS(psc, "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/psc_clusters_compartments.rds")



```


## Epithelial Compartment: Preprocessing and Celltyping
```{r }

# read the original object
psc <- readRDS("/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/psc_clusters_compartments.rds")
#epi <- readRDS("/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/epi_clusters_resct.rds")

#subset for having only the epithelial compartment
epi <- psc[, psc@meta.data$compartment == 'epithelial']
epi <- SCTransform(epi)
DefaultAssay(epi)<-"SCT" 
set.seed(0)

# Preprocessing
epi <- FindVariableFeatures(epi, assay = 'SCT')
#epi <- SCTransform(epi)
#
epi<- RunPCA(epi)
epi <- RunUMAP(epi, dims = 1:30)
epi<-FindNeighbors(epi, dims = 1:30)
epi<-FindClusters(epi, resolution = 1.2)
epi_plot <- DimPlot(epi, group.by = "seurat_clusters", label = T, label.box = T, pt.size = 0.5, repel = T) + NoLegend()
ggsave("epi_clusters.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")
saveRDS(epi, "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/epi_clusters.rds")

#read the trained Smillie dataset object
Smillie_epi_ref <- readRDS("/groups/umcg-weersma/tmp01/datasets/smillie/singlecell_broad_files/Smillie_discovery_cohort/train.Epi.seur.rds")
Smillie_epi_ref <- normalize_and_cluster_sct_method_1(Smillie_epi_ref)
saveRDS(Smillie_epi_ref, "/groups/umcg-weersma/tmp01/datasets/smillie/training_data/train.Epi.seur.sct.rds")

# Perform Azimuth Celltype annotation
DefaultAssay(epi) <- "SCT"
#run azimuth on the non trained dataset
epi <- azimuth_mapping(seurat_object = epi, reference = Smillie_epi_ref)
#saved the new object
saveRDS(epi, '/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/epi_azimuth.rds')
# check how well the classification looks compared to the clusters
p1 <- DimPlot(epi, group.by = "seurat_clusters", label = T, label.box = T, pt.size = 0.5, repel = T) + NoLegend()
p2 <- DimPlot(epi, group.by = "predicted.cell_type.pred", label = T, label.box = T, pt.size = 0.5, repel = T) + NoLegend()
p1 + p2
ggsave("epi_azimuth.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

#ceck the batch effect
epi_lane <- DimPlot(epi, group.by = "lane", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("epi_lane.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

epi_disease <- DimPlot(epi, group.by = "disease", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("epi_disease.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

epi_inflammation <- DimPlot(epi, group.by = "inflammation", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("epi_inflammation.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

epi_state <- DimPlot(epi, group.by = "state", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("epi_state.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

epi_sample <- DimPlot(epi, group.by = "sample", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("epi_sample.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

```

## Epithelial compartment: manually define extra celltypes
```{r }

epi <- readRDS("/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/epi_azimuth.rds")

feature_epi_duox2 <- FeaturePlot(epi, features=c('DUOX2'))
avg_exp_duox2 <- AverageExpression(epi, features = "DUOX2", assays = "RNA")
avg_exp_duox2_table <- data.frame(cluster=colnames(avg_exp_duox2[['RNA']]), expression=as.vector(unlist(avg_exp_duox2[['RNA']])))
#
epi@meta.data$duox2_positive <- NA
epi@meta.data[epi@meta.data[['seurat_clusters']] %in% avg_exp_duox2_table[avg_exp_duox2_table[['expression']] > 4, 'cluster'], 'duox2_positive_4'] <- T
epi@meta.data[epi@meta.data[['seurat_clusters']] %in% avg_exp_duox2_table[avg_exp_duox2_table[['expression']] <= 4, 'cluster'], 'duox2_positive_4'] <- F
epi@meta.data[epi@meta.data[['seurat_clusters']] %in% avg_exp_duox2_table[avg_exp_duox2_table[['expression']] > 2, 'cluster'], 'duox2_positive_2'] <- T
epi@meta.data[epi@meta.data[['seurat_clusters']] %in% avg_exp_duox2_table[avg_exp_duox2_table[['expression']] <= 2, 'cluster'], 'duox2_positive_2'] <- F
epi@meta.data[epi@meta.data[['seurat_clusters']] %in% avg_exp_duox2_table[avg_exp_duox2_table[['expression']] > 3, 'cluster'], 'duox2_positive_3'] <- T
epi@meta.data[epi@meta.data[['seurat_clusters']] %in% avg_exp_duox2_table[avg_exp_duox2_table[['expression']] <= 3, 'cluster'], 'duox2_positive_3'] <- F

epi_duox2_4 <- DimPlot(epi, group.by = 'duox2_positive_4') + scale_color_manual(values=c('gray', 'red'))
epi_duox2_2 <- DimPlot(epi, group.by = 'duox2_positive_2') + scale_color_manual(values=c('gray', 'red'))
epi_duox2_3 <- DimPlot(epi, group.by = 'duox2_positive_3') + scale_color_manual(values=c('gray', 'red'))
epi_dux2_pos <- DimPlot(epi, group.by = 'duox2_positive') + scale_color_manual(values=c('gray', 'red'))

plot_grid(epi_plot, feature_epi_duox2, epi_duox2_2, epi_duox2_3, epi_duox2_4, epi_dux2_pos , nrow = 2, ncol = 3)
ggsave("epi_duox2.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

epi@meta.data$duox2_positive <- NA
epi@meta.data[epi@assays$RNA@data['DUOX2', ] > 1, 'duox2_positive'] <- T
epi@meta.data[epi@assays$RNA@data['DUOX2', ] <= 1, 'duox2_positive'] <- F

epi@meta.data$celltype.final <- epi@meta.data$predicted.cell_type.pred
epi@meta.data[epi@meta.data$predicted.cell_type.pred %in% c('Immature Enterocytes 1',
                                                  'Immature Enterocytes 2',
                                                  'Enterocytes',
                                                  'Best4+ Enterocytes',
                                                  'Enterocyte Progenitors') &
              epi@meta.data$duox2_positive == T, 'celltype.final'] <- 'DUOX2 enterocytes'

epi_duox2 <- DimPlot(epi, group.by = "celltype.final", label = T, label.box = T, pt.size = 0.5, repel = T) + NoLegend()
ggsave("epi_azimuth_duox2.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

```


## Stromal Compartment: Preprocessing and Celltyping
```{r }

# read the original psc object
psc <- readRDS("/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/psc_clusters_compartments.rds")
#stro <- readRDS("/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/stro_clusters_resct.rds")

# subset the object for only the stromal compartment
stro <- psc[, psc@meta.data$compartment == 'stromal']
stro <- SCTransform(stro)
DefaultAssay(stro)<-"SCT"

set.seed(0)
stro <- FindVariableFeatures(stro, assay = 'SCT')

# Preprocessing
stro<- RunPCA(stro)
stro <- RunUMAP(stro, dims = 1:30)
stro <-FindNeighbors(stro, dims = 1:30)
stro <-FindClusters(stro, resolution = 1.2)
stro_plot <- DimPlot(stro, group.by = "seurat_clusters", label = T, label.box = T, pt.size = 0.5, repel = T) + NoLegend()
ggsave("stro_clusters.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")
saveRDS(stro, "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/stro_clusters.rds")

# read the trained Smillie dataset for the stromal compartment
Smillie_fib_ref <- readRDS("groups/umcg-weersma/tmp01/datasets/smillie/singlecell_broad_files/Smillie_discovery_cohort/train.Fib.seur.rds")
Smillie_fib_ref <- normalize_and_cluster_sct_method_1(Smillie_fib_ref)
saveRDS(Smillie_fib_ref, "/groups/umcg-weersma/tmp01/datasets/smillie/training_data/train.Fib.seur.sct.rds")

# Perform automatic azimuth celltype annotation
DefaultAssay(stro) <- "SCT"

#run azimuth for the noin trained 
stro <- azimuth_mapping(seurat_object = stro, reference = Smillie_fib_ref)
#save the final seurat object
saveRDS(stro, "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/stro_azimuth.rds")
p3 <- DimPlot(stro, group.by = "seurat_clusters", label = T, label.box = T, pt.size = 0.5, repel = T) + NoLegend()
p4 <- DimPlot(stro, group.by = "predicted.cell_type.pred", label = T, label.box = T, pt.size = 0.5, repel = T) + NoLegend()
p3 + p4
ggsave("stro_azimuth.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

# check the batch effect
stro_lane <- DimPlot(stro, group.by = "lane", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("stro_lane.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

stro_disease <- DimPlot(stro, group.by = "disease", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("stro_disease.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

stro_inflammation <- DimPlot(stro, group.by = "inflammation", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("stro_inflammation.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

stro_state <- DimPlot(stro, group.by = "state", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("stro_state.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

stro_sample <- DimPlot(stro, group.by = "sample", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("stro_sample.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

```


## Immune Compartment: Preprocessing and Celltyping
```{r }

#read the original psc object
psc <- readRDS("/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/psc_clusters_compartments.rds")

#subset for only the cells in the immune compartment
imm <- psc[, psc@meta.data$compartment == 'immune']

#check the MT content in the immune compartment 
imm_mt <- plot_ncount_vs_mitopct(imm@meta.data)
ggsave("imm_mt.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

imm@meta.data[imm@meta.data[['percent.mt']] < 15,  'threshold_mt_15'] <- T
imm@meta.data[imm@meta.data[['percent.mt']] >= 15, 'threshold_mt_15'] <- F

imm@meta.data[imm@meta.data[['percent.mt']] < 20,  'threshold_mt_20'] <- T
imm@meta.data[imm@meta.data[['percent.mt']] >= 20, 'threshold_mt_20'] <- F


imm_mt_15 <- DimPlot(imm, group.by = 'threshold_mt_15') + scale_color_manual(values=c('gray', 'red'))
imm_mt_20 <- DimPlot(imm, group.by = 'threshold_mt_20') + scale_color_manual(values=c('gray', 'red'))

plot_grid(imm_mt, imm_mt_20, imm_mt_15)
ggsave("imm_mt_15_20.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

# decide for using MT content < 20 
imm <- imm[ , imm@meta.data$percent.mt < 20]
saveRDS(imm, "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/imm_mt20.rds")

#Preprocessing
imm <- SCTransform(imm)
DefaultAssay(imm)<-"SCT"
set.seed(0)
imm <- FindVariableFeatures(imm, assay = 'SCT')

imm<- RunPCA(imm)
imm <- RunUMAP(imm, dims = 1:30)
imm <-FindNeighbors(imm, dims = 1:30)
imm <-FindClusters(imm, resolution = 1.2)
imm_plot <- DimPlot(imm, group.by = "seurat_clusters", label = T, label.box = T, pt.size = 0.5, repel = T) + NoLegend()
ggsave("imm_clusters.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")
saveRDS(imm, "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/imm_azimuth_mt.rds")

#read the trained Smillie immune dataset
Smillie_imm_ref <- readRDS("/data/p307890/Vedo2/batch_2/celltyping/objects/train.Imm.seur.rds")
Smillie_imm_ref <- normalize_and_cluster_sct_method_1(Smillie_imm_ref)
saveRDS(Smillie_imm_ref, "/groups/umcg-weersma/tmp01/datasets/smillie/training_data/train.Imm.seur.sct.rds")
Smillie_imm_ref <- readRDS("/groups/umcg-weersma/tmp01/datasets/smillie/training_data/train.Imm.seur.sct.rds")


# Perform Azimuth automatic celltype annotation
DefaultAssay(imm) <- "SCT"

imm <- azimuth_mapping(seurat_object = imm, reference = Smillie_imm_ref)
#save the obtained seurat object
saveRDS(imm, "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/imm_azimuth.rds")
# check how well the classification looks compared to the clusters
p5 <- DimPlot(imm, group.by = "seurat_clusters", label = T, label.box = T, pt.size = 0.5, repel = T) + NoLegend()
p6 <- DimPlot(imm, group.by = "predicted.cell_type.pred", label = T, label.box = T, pt.size = 0.5, repel = T) + NoLegend()
p5 + p6
ggsave("imm_azimuth.pdf", width = 10, height = 8,path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")


##check batch effect
imm_lane <- DimPlot(imm, group.by = "lane", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("imm_lane.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

imm_disease <- DimPlot(imm, group.by = "disease", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("imm_disease.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

imm_inflammation <- DimPlot(imm, group.by = "inflammation", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("imm_inflammation.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

imm_state <- DimPlot(imm, group.by = "state", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("imm_state.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

imm_sample <- DimPlot(imm, group.by = "sample", label = F, label.box = T, pt.size = 0.5, repel = T) 
ggsave("imm_sample.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")









```


## Immune Compartment: manually define extra celltypes
```{r }



#### IGG
feature_imm_igg <- FeaturePlot(imm, features=c('IGHG1'))
avg_exp_igg <- AverageExpression(imm, features = "IGHG1", assays = "RNA")
avg_exp_igg_table <- data.frame(cluster=colnames(avg_exp_igg[['RNA']]), expression=as.vector(unlist(avg_exp_igg[['RNA']])))


imm@meta.data[imm@meta.data[['seurat_clusters']] %in% avg_exp_igg_table[avg_exp_igg_table[['expression']] >= 90, 'cluster'], 'igg_positive_90'] <- T
imm@meta.data[imm@meta.data[['seurat_clusters']] %in% avg_exp_igg_table[avg_exp_igg_table[['expression']] < 90, 'cluster'], 'igg_positive_90'] <- F

imm_igg_90 <- DimPlot(imm, group.by = 'igg_positive_90') + scale_color_manual(values=c('gray', 'red'))

plot_grid(imm_plot, feature_imm_igg, imm_igg_90)
ggsave("imm_igg_mt.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")




#### IGM
feature_imm_igm <- FeaturePlot(imm, features=c('IGHM'))
avg_exp_igm <- AverageExpression(imm, features = "IGHM", assays = "RNA")
avg_exp_igm_table <- data.frame(cluster=colnames(avg_exp_igm[['RNA']]), expression=as.vector(unlist(avg_exp_igm[['RNA']])))


imm@meta.data[imm@meta.data[['seurat_clusters']] %in% avg_exp_igm_table[avg_exp_igm_table[['expression']] >= 800, 'cluster'], 'igm_positive_800'] <- T
imm@meta.data[imm@meta.data[['seurat_clusters']] %in% avg_exp_igm_table[avg_exp_igm_table[['expression']] < 800, 'cluster'], 'igm_positive_800'] <- F
imm@meta.data[imm@meta.data[['seurat_clusters']] %in% avg_exp_igm_table[avg_exp_igm_table[['expression']] >= 700, 'cluster'], 'igm_positive_700'] <- T
imm@meta.data[imm@meta.data[['seurat_clusters']] %in% avg_exp_igm_table[avg_exp_igm_table[['expression']] < 700, 'cluster'], 'igm_positive_700'] <- F


imm_igm_800 <- DimPlot(imm, group.by = 'igm_positive_800') + scale_color_manual(values=c('gray', 'red'))
imm_igm_700 <- DimPlot(imm, group.by = 'igm_positive_700') + scale_color_manual(values=c('gray', 'red'))


plot_grid(imm_plot, feature_imm_igm, imm_igm_700, imm_igm_800)
ggsave("imm_igm_mt.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")



#### IGA
imm@meta.data[imm@meta.data[['seurat_clusters']] %in% avg_exp_iga_table[avg_exp_iga_table[['expression']] >= 120, 'cluster'], 'iga_positive_120'] <- T
imm@meta.data[imm@meta.data[['seurat_clusters']] %in% avg_exp_iga_table[avg_exp_iga_table[['expression']] < 120, 'cluster'], 'iga_positive_120'] <- F
imm@meta.data[imm@meta.data[['seurat_clusters']] %in% avg_exp_iga_table[avg_exp_iga_table[['expression']] >= 115, 'cluster'], 'iga_positive_115'] <- T
imm@meta.data[imm@meta.data[['seurat_clusters']] %in% avg_exp_iga_table[avg_exp_iga_table[['expression']] < 115, 'cluster'], 'iga_positive_115'] <- F

imm_iga_120 <- DimPlot(imm, group.by = 'iga_positive_120') + scale_color_manual(values=c('gray', 'red'))
imm_iga_115 <- DimPlot(imm, group.by = 'iga_positive_115') + scale_color_manual(values=c('gray', 'red'))

plot_grid(feature_imm_iga, imm_iga_115, imm_iga_120)
ggsave("imm_iga_mt.pdf", width = 10, height = 8, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

```

## Immune Compartment: Define Plasma Compartment
```{r }

imm@meta.data$plasma_compartment <- 'not_plasma'
imm@meta.data[imm@meta.data$predicted.cell_type.pred == 'Plasma', 'plasma_compartment'] <- 'Ig_negative'
imm@meta.data[imm@meta.data$iga_positive_115 == T, 'plasma_compartment'] <- 'IgA'
imm@meta.data[imm@meta.data$igg_positive_90 == T, 'plasma_compartment'] <- 'IgG'
imm@meta.data[imm@meta.data$igm_positive_800 == T, 'plasma_compartment'] <- 'IgM'

compart_plasma <- DimPlot(imm, group.by = 'plasma_compartment') + scale_color_manual(values=c('firebrick', 'dodgerblue4', 'green4', 'black','grey'))
plot_grid(imm_plot, compart_plasma)

ggsave("imm_plasma_compart.pdf", width = 20, height = 16, path = "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/plots")

saveRDS(imm, "/groups/umcg-weersma/tmp01/singelcell/PSCproject/Processed_data/results_07_2022/objects/imm_azimuth_plasma_compartments.rds")

```


## Create input for Cell Abundances analysis
```{r }

#Read the last version for compartment of the seurat object
imm <- readRDS("/Users/paolapibiri/Desktop/PSC/celltype_annotation/new_results /objects/imm_azimuth_with_plasma.rds")
epi <- readRDS("/Users/paolapibiri/Desktop/PSC/celltype_annotation/new_results /objects/epi_azimuth_duox2.rds")
stro <- readRDS("/Users/paolapibiri/Desktop/PSC/celltype_annotation/new_results /objects/stro_azimuth.rds") 


# extract counts to matrix for each compartment
imm@meta.data$condition <- paste0(imm@meta.data$sample, '_', imm@meta.data$disease, '_', imm@meta.data$inflammation)
imm_counts=as.data.frame.matrix(table(imm@meta.data[, c('condition', 'predicted.cell_type.pred')]))
imm_counts['condition'] <- rownames(imm_counts)
imm_counts <- imm_counts[, c('condition', setdiff(colnames(imm_counts), 'condition'))]
#

epi@meta.data$condition <- paste0(epi@meta.data$sample, '_', epi@meta.data$disease, '_', epi@meta.data$inflammation)
epi_counts=as.data.frame.matrix(table(epi@meta.data[, c('condition', 'celltype.final')]))
epi_counts['condition'] <- rownames(epi_counts)
epi_counts <- epi_counts[, c('condition', setdiff(colnames(epi_counts), 'condition'))]
#
stro@meta.data$condition <- paste0(stro@meta.data$sample, '_', stro@meta.data$disease, '_', stro@meta.data$inflammation)
stro_counts=as.data.frame.matrix(table(stro@meta.data[, c('condition', 'predicted.cell_type.pred')]))
stro_counts['condition'] <- rownames(stro_counts)
stro_counts <- stro_counts[, c('condition', setdiff(colnames(stro_counts), 'condition'))]

#save the dataframe in csv file 
write.csv(imm_counts, '/Users/paolapibiri/Desktop/psc_imm_count_07.csv')
write.csv(epi_counts, '/Users/paolapibiri/Desktop/psc_epi_count_07.csv')
write.csv(stro_counts, '/Users/paolapibiri/Desktop/psc_stro_count_07.csv')

##merge the 3 dataframes together
imm_epi <- merge(imm_counts, epi_counts, by= 'condition')
imm_epi_stro <-  merge(imm_epi, stro_counts, by= 'condition')

write.csv(imm_epi_stro, '/Users/paolapibiri/Desktop/psc_counts_07.csv')


epi_stro_imm_2 <- add_data_from_rownames(imm_epi_stro, column_names = c('sample', 'disease', 'inflammation'))
epi_stro_imm_2$condition <- NULL
epi_stro_imm_2$sample <- NULL
write.csv(epi_stro_imm_2, '/Users/paolapibiri/Desktop/psc_counts_07.csv')

### filter Inflamed
epi_stro_imm_2_inflamed <- epi_stro_imm_2[epi_stro_imm_2$inflammation == 'I', ]
epi_stro_imm_2_inflamed[['inflammation']] <- NULL

write.csv(epi_stro_imm_2_inflamed, '/Users/paolapibiri/Desktop/psc_counts_I.csv')

### filter Non-nflamed
epi_stro_imm_2_noninflamed <- epi_stro_imm_2[epi_stro_imm_2$inflammation == 'NI', ]
epi_stro_imm_2_noninflamed[['inflammation']] <- NULL

write.csv(epi_stro_imm_2_noninflamed, '/Users/paolapibiri/Desktop/psc_counts_NI.csv')

### filter UC
epi_stro_imm_2_uc <- epi_stro_imm_2[epi_stro_imm_2$disease == 'UC', ]
epi_stro_imm_2_uc[['disease']] <- NULL

write.csv(epi_stro_imm_2_uc, '/Users/paolapibiri/Desktop/psc_counts_UC.csv')

### filter PSC
epi_stro_imm_2_psc <- epi_stro_imm_2[epi_stro_imm_2$disease == 'PSC', ]
epi_stro_imm_2_psc[['disease']] <- NULL

write.csv(epi_stro_imm_2_psc, '/Users/paolapibiri/Desktop/psc_counts_PSC.csv')

##filter for HC and UC I
psc <- read.csv('/Users/paolapibiri/Desktop/PSC/compositonal_analysis/new_results/psc_scCODA_inputs/psc_counts_07.csv', check.names = F)
psc$X <- NULL

epi_stro_imm_hc_uc_i <- psc[psc$disease == 'HC' | (psc$disease == 'UC' & psc$inflammation == 'I'), ]
epi_stro_imm_hc_uc_i[['status']] <- paste(epi_stro_imm_hc_uc_i$disease, epi_stro_imm_hc_uc_i$inflammation, sep = '-')
epi_stro_imm_hc_uc_i$inflammation <- NULL
epi_stro_imm_hc_uc_i$disease <- NULL

write.csv(epi_stro_imm_hc_uc_i, '/Users/paolapibiri/Desktop/psc_counts_UCI_HC.csv',  row.names = F)

##filter for HC and PSC I
psc <- read.csv('/Users/paolapibiri/Desktop/PSC/compositonal_analysis/new_results/psc_scCODA_inputs/psc_counts_07.csv', check.names = F)
psc$X <- NULL

epi_stro_imm_hc_psc_i <- psc[psc$disease == 'HC' | (psc$disease == 'PSC' & psc$inflammation == 'I'), ]
epi_stro_imm_hc_psc_i[['status']] <- paste(epi_stro_imm_hc_psc_i$disease, epi_stro_imm_hc_psc_i$inflammation, sep = '-')
epi_stro_imm_hc_psc_i$inflammation <- NULL
epi_stro_imm_hc_psc_i$disease <- NULL

write.csv(epi_stro_imm_hc_psc_i, '/Users/paolapibiri/Desktop/psc_counts_PSCI_HC.csv',  row.names = F)


```

